Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-05 dsn-2018/bare_conf.tex
DSN changes

Cryptographic hash functions form the core of many protocols which
trust that cryptographic hash functions preserve certain basic
properties. Cryptograpic hash functions see widespread
use for file integrity checks to verify long term storage of
data, as cache invalidation techniques, and as a building block in network
protocols such as Kerberos and TLS.  The classes of functions suitable
for these use cases necessarily have
strong guarantees about properties of its members. 
An ideal cryptographic hash has five main features: determinism - the
same message always results in the same hash; performability -
computing a hash for a message should not be algorithmically
difficult; one-way - it should be infeasible to generate a message
from its hash except by trying all possible messages; sensitive to
initial conditions - small changes to a message should change the hash
value to a degree that the new hash value appears uncorrelated with
the old hash value; unique digests - it should be infeasible to find
two different messages with the same hash value.

The most important aspect of these can be formally defined as three
properties which hash functions must have to be considered cryptographically secure:
\begin{itemize}
    \item \textbf{Preimage Resistance}: It should be computationally hard to find
        the inverse of a hash function.
    \item \textbf{Second Preimage Resistance}: Given an input block, it should be
        computationally hard to find a second block which hashes to the
        same value as the first block.
    \item \textbf{Collision Resistance}: It should be computationally hard to find two
        blocks which hash to the same value.
\end{itemize}
Note that while a second preimage is necessarily a collision, it is also a
stronger result with higher use for an attacker.

@@

From a cryptographic perspective, the existence of a feasible collision attack
breaks the requisite properties, thus deprecating the function's use
for trustworthy applications. In many practical instances, however,
communities continue to use deprecated hash functions in a widespread fashion.
One such example of widespread use post-deprecation is the continued use by Git of SHA-1, despite
M. Stevens et al. publishing a full SHA-1 collision in February of 2017
\cite{cryptoeprint:2017:190}. This attack had a bound of $2^{63.1}$
\emdash faster than the theoretical $2^{80}$ but still requiring significant compute
resources to find.

We seek to address the problem of evaluating the use cases for a collision,
to determine whether or not a collision affects a particular system. This
paper contributes several techniques to logical cryptanalysis and uses these
techniques to derive metrics of the trustworthiness of hash functions. We claim
the following results are novel in the field:
\begin{itemize}
    \item We define the neighborhood of a class of collisions and show that
        it is frequently non-empty.
    \item We discuss techniques for finding useful collisions within a
        class of collisions using logical cryptanalysis.
    \item We propose additional techniques for building new classes of
        collisions and show that collisions in MD4 can be found inductively.
    \item We propose five universal metrics for evaluating the utility of
        a class of collisions.
\end{itemize}
We propose that collision classes of high utility are closer to a second
preimage attack in the amount of flexibility they provide to an attacker,
whereas classes of low utility may not affect all systems which use a
particular hash function.

@@

@@

@@

@@

We define a relation among families of collisions across rounds as follows.
Let $F_1$ and $F_2$ be two different families of collisions. Then we say that
$F_1$ and $F_2$ are \textit{similar}, and notate it $F_1 \lesssim F_2$, if
$R(F_1) \leq R(F_2)$ and $F_1 \subseteq F_2$. Note that, when
$R(F_1) = R(F_2)$, this is equivalent to saying that $F_2$ is in some
expansion neighborhood of $F_1$. Further, when $R(F_1) < R(F_2)$ and
$F_1 = F_2$, then we say that $F_2$ is the \textit{trivial extension} of
$F_1$.

We claim that the following statements are true for MD4:
\begin{enumerate}
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F' \lesssim F$ and $R(F') + 4 = R(F)$.
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F \lesssim F'$.
\end{enumerate}

@@

We share the general tips for working with models of MD4 from a SAT solver
performance perspective, without justification for results. Note that these
results may be limited in usefulness and may not be uniformly applicable.
\begin{itemize}
    \item Unless searching for a second preimage or chosen prefix with
        sufficient number of bits ($\geq 128$), we find models perform better
        leaving the initial state unconstrained.
    \item Adding ASCII block constraints significantly and negatively impact
        the performance of finding a collision.
    \item Keeping a small search radius on neighborhoods involving families
        of collision classes is important for good performance. However,
        once sufficient witnesses have been found across rounds, missing
        families within the round can be found by various filling-in
        techniques through applying the $\lesssim$ relation.
\end{itemize}

>>>>

Cryptographic hash functions form the core of many protocols which
trust that cryptographic hash functions preserve certain basic
properties. Cryptograpic hash functions see widespread
use for file integrity checks to verify long term storage of
data, as cache invalidation techniques, and as a building block in network
protocols such as Kerberos and TLS.  The classes of functions suitable
for these use cases necessarily have
strong guarantees about properties of its members. 
An ideal cryptographic hash has five main features: determinism - the
same message always results in the same hash; performability -
computing a hash for a message should not be algorithmically
difficult; non-reversability - it should be infeasible to generate a message
from its hash except by trying all possible messages; sensitivity to
initial conditions - small changes to a message should change the hash
value to a degree that the new hash value appears uncorrelated with
the old hash value; unique digests - it should be infeasible to find
two different messages with the same hash value.

The most important aspect of these can be formally defined under three
properties which hash function implementations must have to be
considered cryptographically secure in practice:
\begin{itemize}
    \item \textbf{Preimage Resistance}: It should be computationally hard to find
        the inverse of a hash function.
    \item \textbf{Second Preimage Resistance}: Given an input block, it should be
        computationally hard to find a second block which hashes to the
        same value as the first block.
    \item \textbf{Collision Resistance}: It should be computationally hard to find two
        blocks which hash to the same value.
\end{itemize}
Note that while a second preimage is necessarily a collision, it is also a
stronger result with higher use for an attacker.

@@

From a cryptographic perspective, the existence of a feasible collision attack
breaks the requisite properties, thus deprecating the function's use
for trustworthy applications. In many practical instances, however,
communities continue to use deprecated hash functions in a widespread fashion.
One such example of widespread use post-deprecation is the continued use by Git of SHA-1, despite
M. Stevens et al. publishing a full SHA-1 collision in February of 2017
\cite{cryptoeprint:2017:190}. This attack had a bound of $2^{63.1}$
--- faster than the theoretical $2^{80}$ but still requiring significant compute
resources to find.

We seek to address the problem of evaluating the use cases for a collision,
to determine whether or not a collision affects a particular system.
The central thesis of this research is that logical cryptanalysis
allows us to formalize the structure inherent to the underlying hash
function such that new collisions can be found within reasonable time
on commodity hardware.  We assert that when such structure exists, it
is \textbf{imperitive to discontinue use of such hash functions as
  they are fundamentally untrustworthy for critical applications.}  An
example of such structure showing the unit distance neighborhood
(defined in Section \ref{Sec:Terminology}) of the known collisions
from Dobbertin and Kasselman is presented in Figure \ref{Fig:Graph-Intro}.
This paper contributes several techniques to logical cryptanalysis and uses these
techniques to derive metrics of the trustworthiness for hash functions. We claim
the following results are novel in the field:
\begin{itemize}
    \item We define the neighborhood of a class of collisions and show that
        it is frequently non-empty.
    \item We discuss techniques for finding useful collisions within a
        class of collisions using logical cryptanalysis.
    \item We propose additional techniques for building new classes of
        collisions and show that collisions in MD4 can be found inductively.
    \item We propose five universal metrics for evaluating the utility of
        a class of collisions.
\end{itemize}
We propose that collision classes of high utility are closer to a second
preimage attack in the amount of flexibility they provide to an attacker,
whereas classes of low utility may not affect all systems which use a
particular hash function.

@@

We introduce this distance function as a heuristic measure of the
sensitivity of the hash function to small changes in the input.

@@

When the neighborhood defined by $\delta = 1$ is frequently non-empty
we assert that sufficient structure exists in the underlying hash
function such that logical cryptanalysis via 3-CNF-SAT becomes
feasible.  This condition is thus sufficient to render the underlying
hash function \textbf{untrustworthy}.  Conversely it is a necessary
condition (but possibly insufficient) that a trustworthy hash function
has an empty neighborhood defined by $\delta = 1$.

Furthermore this condition extends to $\delta = d$ for some small
value of $d$ as determined by computational resources such that $d$
becomes a trustworthiness metric under constraints of current
3-CNF-SAT solving capabilities within reasonable time bounds.

@@

We find justification for the distance metric $\Delta$ in the existing
literature due to collisions by  X. Wang \cite{cryptoeprint:2004:199} and
M. Schl{\"a}ffer \cite{Schlaffer2006}. Their respective families are:

\begin{align*}
F(C_{Wang}) &=& (1, 2, 3, 5, 6, 7, 8, 9, 10, 11, 12, \textbf{13}, 15, 16, 19, 20, 35, 36)\\
F(C_{Schlaffer}) &=& (1, 2, 3, 5, 6, 7, 8, 9, 10, 11, 12, 15, 16, 19, 20, 35, 36)
\end{align*}

@@

We define a relation among families of collisions across rounds as follows.
Let $F_1$ and $F_2$ be two different families of collisions. Then we say that
$F_1$ and $F_2$ are \textit{similar}, and notate it $F_1 \precsim F_2$, if
$R(F_1) \leq R(F_2)$ and $F_1 \subseteq F_2$. Note that, when
$R(F_1) = R(F_2)$, this is equivalent to saying that $F_2$ is in some
expansion neighborhood of $F_1$. Further, when $R(F_1) < R(F_2)$ and
$F_1 = F_2$, then we say that $F_2$ is the \textit{trivial extension} of
$F_1$.

We claim that the following statements are true for MD4:
\begin{enumerate}
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F' \precsim F$ and $R(F') + 4 = R(F)$.
    \item For every $F \in \mathcal{F}$, there exists $F'$ such that
        $F \precsim F'$.
\end{enumerate}

@@

We share the general tips for working with models of MD4 from a SAT solver
performance perspective, without justification for results. Note that these
results may be limited in usefulness and may not be uniformly applicable.
\begin{itemize}
    \item Unless searching for a second preimage or chosen prefix with
        sufficient number of bits ($\geq 128$), we find models perform better
        leaving the initial state unconstrained.
    \item Adding ASCII block constraints significantly and negatively impact
        the performance of finding a collision.
    \item Keeping a small search radius on neighborhoods involving families
        of collision classes is important for good performance. However,
        once sufficient witnesses have been found across rounds, missing
        families within the round can be found by various filling-in
        techniques through applying the $\precsim$ relation.
\end{itemize}
